Galois Slicing as Differentiation

### 2025-07-11 main/models.tex
Tweaks.
Typo.
Obsolete.

\begin{enumerate}
\item The category $\LatGal$ has bounded lattices as objects and
  Galois connections as morphisms, with the right adjoint going in the
  ``forward'' direction. The category $\Fam(\LatGal)$ is our preferred
  model for total functions with Galois sliicing. We explore some
  specific examples in this category in \secref{semantic-gps}.
\item The category $\MeetSLat$ has meet-semilattices with top as
  objects and monotone finite meet preserving functions as
  morphisms. The category $\Fam(\MeetSLat)$ provides a model of
  total functions with ``forward derivatives'' only.
\item The category $\JoinSLat$ has join-semilattices with bottom as
  objects and monotone finite join preserving functions as
  morphisms. The category $\Fam(\JoinSLat^\op)$ provides a model of
  total functions with ``backwards derivatives'' only.
\end{enumerate}

@@

  \noindent satisfying

  \vspace{-3mm}
  \begin{minipage}[t]{0.45\textwidth}
    \begin{center}
      \begin{salign*}
        \biproj_X \comp \biinj_X &= \id_X \\
        \biproj_Y \comp \biinj_X &= \zero_{X,Y}
      \end{salign*}
    \end{center}
  \end{minipage}%
  \begin{minipage}[t]{0.45\textwidth}
    \begin{center}
      \begin{salign*}
        \biproj_Y \comp \biinj_Y &= \id_Y \\
        \biproj_X \comp \biinj_Y &= \zero_{Y,X}
      \end{salign*}
    \end{center}
  \end{minipage}

@@

The operation of adding a new bottom element to a bounded lattice
forms part of a monad $L$ on $\LatGal$. This monad extends to a
(strong) monad $\Lift$ on $\Fam(\LatGal)$ with
$\Lift(X, \partial X) = (X, L \circ \partial X)$. The monad $\Lift$
does not affect the points of the original object, but adds a new
minimum approximation. This monad allows us to construct morphisms in
$\Fam(\LatGal)$ that track information about how values are used
similar to the examples of cm and stable functions we saw in
\secref{approx-as-tangents}.

@@

Let $\Bool = \Disc(\{\mathsf{tt},\mathsf{ff}\})$ be the
(unapproximated) embedding of the booleans and
$\mathrm{or} : \Bool \times \Bool \to \Bool$ be the (unapproximated)
boolean or function. The, using a Moggi-style let notation
\cite{notions-of-computation} for morphisms constructed using the
Monad structure of $\Lift$, we can reproduce the functions
$\mathrm{strictOr}$ and $\mathrm{shortCircuitOr}$ functions from
\exref{strict-short-circuit} (we also assume an if-then-else operation
on booleans, definable from the fact that $\Fam(\LatGal)$ has
coproducts and $\Disc$ preserves them). Both of these expressions
define morphisms $\Lift(\Bool) \times \Lift(\Bool) \to \Lift(\Bool)$
in $\Fam(\LatGal)$:
\begin{displaymath}
  \begin{array}{lcl}
    \mathrm{strictOr}(x,y)&=&\mathrm{let}\,b_1 \Leftarrow x\,\mathrm{in}\,\mathrm{let}\,b_2 \Leftarrow y\,\mathrm{in}\,\eta(\mathrm{or}(b_1,b_2)) \\
    \mathrm{shortCircuitOr}(x,y)&=&\mathrm{let}\,b_1 \Leftarrow x\,\mathrm{in}\,\mathrm{if}\,b_1\,\mathrm{then}\,\eta(\mathsf{tt})\,\mathrm{else}\,y
  \end{array}
\end{displaymath}
Examining the morphisms so defined in $\Fam(\LatGal)$, we can see
that, in the $\Set$ component, they are both exactly the normal
boolean-or operation. However, they have different approximation
behaviour, reflecting the different ways that they examine their
inputs. Let us write $\top,\bot$ for the elements of the approximation
lattice at each point of $\Lift(\Bool)$, then applying the reverse
derivative at the point $(\mathsf{tt},\mathsf{tt})$ to the tangent
$\top$ reveals which of the inputs contributed to the output for each
function:
\begin{displaymath}
  \begin{array}{lcl}
    (\partial \mathrm{strictOr})_r(\mathsf{tt},\mathsf{tt})(\top) &=& (\top, \top)  \\
    (\partial \mathrm{shortCircuitOr})_r(\mathsf{tt},\mathsf{tt})(\top) &=& (\top, \bot)
  \end{array}
\end{displaymath}
In comparison to the categories $\CM$ and $\Lposet$ from
\secref{approx-as-tangents}, we have retained the usage information in
the forward and reverse tangents, but we also accurately model
totality of the functions. That is, the constantly $\bot$ function is
also present in both $\CM$ and $\Lposet$, but is not expressible in
$\Fam(\LatGal)$.

@@

\begin{remark}
  \label{rem:further-structure}
  These examples highlight a potential criticism of $\Fam(\LatGal)$ as
  a category for modelling \GPS. For $\mathrm{shortCircuitOr}$, we had
  $(\partial \mathrm{shortCircuitOr})_r(\mathsf{tt},\mathsf{tt})(\top)
  = (\top, \bot)$, indicating that the second argument was not needed
  for computing the output. However, there is no way, in
  $\Fam(\LatGal)$, of turning this into a rigorous statement that the
  $\Set$-component of this morphism does not actually depend on its
  second argument. We conjecture that this can be rectified by
  requiring some kind of additional structure on each object
  $(X, \partial X)$ consisting of a map
  $\Pi_{x : X}.\partial X(x) \to \mathcal{P}(X)$, where $\mathcal{P}$
  is the powerset, which identifies for each $x$ which elements are
  indistinguishable from $x$ at this level of approximation. One would
  also presumably have to require additional conditions for this to
  respect the lattice structure and be preserved by
  morphisms\footnote{This additional structure is reminiscent of the
    additional structure on \emph{directed containers} defined by
    \citet{ahman-chapman-uustalu2012}. They require a map
    $\Pi_{x : X}.\partial X(x) \to X$ picking out a specific $X$
    ``jumped to'' by some change $\delta x : \partial X(x)$ at $x$. In
    our proposed setup, we follow the approximation theme of \GPS by
    having a \emph{set} of things that could be taken instead of the
    original $x$. Objects of $\Fam(\LatGal)$ are ``directed'' in this
    sense because the approximation map can pick out the element of
    the original poset that was approximating $x$.}.
\end{remark}

@@

This example shows that we have to be careful about how we choose the
interpretation of approximable sets in $\Fam(\LatGal)$, and again
highlights the point we made in \remref{further-structure} that
perhaps $\Fam(\LatGal)$ does not have quite enough structure to
determine ``sensible'' approximation information. On the other hand,
the use of the $T$ monad does have the advantage that the
approximation lattices built from discrete sets, products, and
coproducts, are always Boolean lattices, meaning that we can take
complements of usage information. The ability to take boolean
complements of approximations has been used by \citet{perera22} to
compute \emph{related} outputs, as we discuss in
\secref{related-work:galois-slicing}.

@@

So far, the approximation lattices we have looked at in
$\Fam(\LatGal)$ have only consisted of those constructed from finite
products and lifting, and only track binary usage/non-usage
information. \exref{intervals-and-maxima-elements} shows how we can go
beyond this to get more ``quantitative'' approximation
information. Let the object of reals with interval approximations in
$\Fam(\LatGal)$ be
$\mathbb{R}_{\mathit{intv}} = (\mathbb{R}, \lambda x.\,\{[l,u] \mid l
\leq x \leq u\} \cup \{\bot\})$ where the lattices of intervals are
reverse ordered by inclusion with $\bot$ at the bottom. Then,
following the examples in \exref{intervals-and-maxima-elements} and
\exref{stable-functions}, we can define addition, negation, and
scaling by $r \geq 0$:
\begin{displaymath}
  \begin{array}{lcl}
    \mathrm{add}&=&(\begin{array}[t]{@{}l}
      \lambda (x_1,x_2).\,x_1+x_2, \\
      \lambda (x_1,x_2)\,([l_1,u_1],[l_2,u_2]).\,[(l_1+x_2) \sqcap (l_2+x_1),(u_1+x_2) \sqcup (u_2+x_1)] \\
      \lambda (x_1,x_2)\,[l,u].\,([l-x_2,u-x_2],[l-x_1,u-x_1]))
    \end{array}\\
    \mathrm{neg}&=&(\lambda x.\,-x, \lambda x\,[l,u].\,[-u,-l], \lambda x\,[l,u].\,[-u,-l]) \\
    \mathrm{scale}(r)&=&(\lambda x.\,rx, \lambda x\,[l,u].\,[rl,ru], \lambda x\,[l,u].\,\mathrm{if}\,r=0\,\mathrm{then}\,\bot\,\mathrm{else}\,[\frac{l}{r},\frac{u}{r}])
  \end{array}
\end{displaymath}
(we only define the forward and backward maps on intervals, their
behaviour on $\bot$ is determined.)  Scaling by negative numbers is
also possible with swapping of lower and upper bounds, as is
multiplication.

@@

Comparing this conjecture to the previous one, we can see that the
difference between $\CM$, $\Lposet$ and $\Fam(\MeetSLat)$,
$\Fam(\LatGal)$ is that the latter have a separation of points from
tangents, somewhat analogous to the situation with manifolds. Also
note that, if this conjecture holds, then contrary to the
$\Fam(\FinVect)$ representation of manifolds and differentiable maps,
we do not throw away information about higher derivatives. It is
retained in the order structure of the tangent fibres.

>>>>

\begin{enumerate}
\item The category $\LatGal$ has bounded lattices as objects and
  Galois connections as morphisms, with the right adjoint going in the
  ``forward'' direction. The category $\Fam(\LatGal)$ is our preferred
  model for total functions with Galois slicing. We explore some
  specific examples in this category in \secref{semantic-gps}.
\item The category $\MeetSLat$ has meet-semilattices with top as
  objects and monotone finite meet preserving functions as
  morphisms. The category $\Fam(\MeetSLat)$ provides a model of
  total functions with ``forward derivatives'' only.
\item The category $\JoinSLat$ has join-semilattices with bottom as
  objects and monotone finite join preserving functions as
  morphisms. The category $\Fam(\JoinSLat^\op)$ provides a model of
  total functions with ``backwards derivatives'' only.
\end{enumerate}

@@

  \vspace{-1mm}
  \noindent satisfying

  \vspace{-5mm}
  \begin{minipage}[t]{0.45\textwidth}
    \begin{center}
      \begin{salign*}
        \biproj_X \comp \biinj_X &= \id_X \\
        \biproj_Y \comp \biinj_X &= \zero_{X,Y}
      \end{salign*}
    \end{center}
  \end{minipage}%
  \begin{minipage}[t]{0.45\textwidth}
    \begin{center}
      \begin{salign*}
        \biproj_Y \comp \biinj_Y &= \id_Y \\
        \biproj_X \comp \biinj_Y &= \zero_{Y,X}
      \end{salign*}
    \end{center}
  \end{minipage}

@@

The operation of adding a new bottom element to a bounded lattice
forms part of a monad $L$ on $\LatGal$. This monad extends to a
(strong) monad $\Lift$ on $\Fam(\LatGal)$ with
$\Lift(X, \partial X) = (X, L \circ \partial X)$. The monad $\Lift$
does not affect the points of the original object, but adds a new
minimum approximation.

@@

Let $\Bool = \Disc(\{\mathsf{tt},\mathsf{ff}\})$ be the
(unapproximated) embedding of the booleans and
$\mathrm{or} : \Bool \times \Bool \to \Bool$ be the (unapproximated)
boolean OR function. Using a Moggi-style let notation
\cite{notions-of-computation} for morphisms constructed using the
Monad structure of $\Lift$, we can reproduce the functions
$\mathrm{strictOr}$ and $\mathrm{shortCircuitOr}$ functions from
\exref{strict-short-circuit} (we also assume an if-then-else operation
on booleans, definable from the fact that $\Fam(\LatGal)$ has
coproducts and $\Disc$ preserves them). Both of these expressions
define morphisms $\Lift(\Bool) \times \Lift(\Bool) \to \Lift(\Bool)$
in $\Fam(\LatGal)$:
\begin{displaymath}
  \begin{array}{lcl}
    \mathrm{strictOr}(x,y)&=&\mathrm{let}\,b_1 \Leftarrow x\,\mathrm{in}\,\mathrm{let}\,b_2 \Leftarrow y\,\mathrm{in}\,\eta(\mathrm{or}(b_1,b_2)) \\
    \mathrm{shortCircuitOr}(x,y)&=&\mathrm{let}\,b_1 \Leftarrow x\,\mathrm{in}\,\mathrm{if}\,b_1\,\mathrm{then}\,\eta(\mathsf{tt})\,\mathrm{else}\,y
  \end{array}
\end{displaymath}
Examining the morphisms so defined in $\Fam(\LatGal)$, we can see
that, in the $\Set$ component, they are both exactly the normal
boolean-or operation. However, they have different approximation
behaviour, reflecting the different ways that they examine their
inputs. Let us write $\top,\bot$ for the elements of the approximation
lattice at each point of $\Lift(\Bool)$, then applying the reverse
derivative at $(\mathsf{tt},\mathsf{tt})$ to the tangent $\top$
reveals which of the inputs contributed to the output for each
function:
\begin{displaymath}
  \begin{array}{lcl}
    (\partial \mathrm{strictOr})_r(\mathsf{tt},\mathsf{tt})(\top) &=& (\top, \top)  \\
    (\partial \mathrm{shortCircuitOr})_r(\mathsf{tt},\mathsf{tt})(\top) &=& (\top, \bot)
  \end{array}
\end{displaymath}
In comparison to the categories $\CM$ and $\Lposet$ from
\secref{approx-as-tangents}, we have retained the usage information in
the forward and reverse tangents, but we also accurately model
totality of the functions. That is, the constantly $\bot$ function is
also present in both $\CM$ and $\Lposet$, but is not expressible in
$\Fam(\LatGal)$.

@@

\begin{remark}
  \label{rem:further-structure}
  These examples highlight a potential criticism of $\Fam(\LatGal)$ as
  a category for modelling \GPS. For $\mathrm{shortCircuitOr}$, we had
  $(\partial \mathrm{shortCircuitOr})_r(\mathsf{tt},\mathsf{tt})(\top)
  = (\top, \bot)$, indicating that the second argument was not needed
  for computing the output. However, there is no way, in
  $\Fam(\LatGal)$, of turning this into a rigorous statement that the
  $\Set$-component of this morphism does not actually depend on its
  second argument. We conjecture that this can be rectified by
  requiring some kind of additional structure on each object
  $(X, \partial X)$ consisting of a map
  $\Pi_{x : X}.\partial X(x) \to \mathcal{P}(X)$, where $\mathcal{P}$
  is the powerset, which identifies for each $x$ which elements are
  indistinguishable from $x$ at this level of approximation. One would
  also presumably have to require additional conditions for this to
  respect the lattice structure and be preserved by
  morphisms\footnote{This additional structure is reminiscent of the
    additional structure on \emph{directed containers} defined by
    \citet{ahman-chapman-uustalu2012}. They require a map
    $\Pi_{x : X}.\partial X(x) \to X$ picking out a specific $X$
    ``jumped to'' by some change $\delta x : \partial X(x)$ at $x$. In
    our proposed setup, we follow the approximation theme of \GPS by
    having a \emph{set} of things that could be used to replace the
    original $x$. We observe that objects of $\Fam(\LatGal)$ arising
    from $\Lposet$ are ``directed'' in the
    \citet{ahman-chapman-uustalu2012} sense because the map can pick
    out the element of the original poset that was approximating
    $x$.}.
\end{remark}

@@

This example shows that we have to be careful about how we choose the
interpretation of approximable sets in $\Fam(\LatGal)$, and again
highlights the point we made in \remref{further-structure} that
perhaps $\Fam(\LatGal)$ does not have quite enough structure to
determine ``sensible'' approximation information. On the other hand,
the use of the $T$ monad does have the advantage that the
approximation lattices built from discrete sets, products, and
coproducts, are always Boolean lattices, meaning that we can take
complements of usage information. The ability to take complements of
approximations has been used by \citet{perera22} to compute
\emph{related} outputs, as we discuss in
\secref{related-work:galois-slicing}.

@@

So far, the approximation lattices we have looked at in
$\Fam(\LatGal)$ have only consisted of those constructed from finite
products and lifting, and only track binary usage/non-usage
information. \exref{intervals-and-maxima-elements} shows how we can go
beyond this to get more ``quantitative'' approximation
information. Let the object of reals with interval approximations in
$\Fam(\LatGal)$ be
$\mathbb{R}_{\mathit{intv}} = (\mathbb{R}, \lambda x.\,\{[l,u] \mid l
\leq x \leq u\} \cup \{\bot\})$ where the lattices of intervals are
reverse ordered by inclusion with $\bot$ at the bottom. Then,
following the examples in \exref{intervals-and-maxima-elements} and
\exref{stable-functions}, we can define addition, negation, and
scaling by $r \geq 0$:
\begin{displaymath}
  \begin{array}{lcl}
    \mathrm{add}&=&(\begin{array}[t]{@{}l}
      \lambda (x_1,x_2).\,x_1+x_2, \\
      \lambda (x_1,x_2)\,([l_1,u_1],[l_2,u_2]).\,[(l_1+x_2) \sqcap (l_2+x_1),(u_1+x_2) \sqcup (u_2+x_1)] \\
      \lambda (x_1,x_2)\,[l,u].\,([l-x_2,u-x_2],[l-x_1,u-x_1]))
    \end{array}\\
    \mathrm{neg}&=&(\lambda x.\,{-}x, \lambda x\,[l,u].\,[-u,-l], \lambda x\,[l,u].\,[-u,-l]) \\
    \mathrm{scale}(r)&=&(\lambda x.\,rx, \lambda x\,[l,u].\,[rl,ru], \lambda x\,[l,u].\,\mathrm{if}\,r=0\,\mathrm{then}\,\bot\,\mathrm{else}\,[\frac{l}{r},\frac{u}{r}])
  \end{array}
\end{displaymath}
(we only define the forward and backward maps on intervals, their
behaviour on $\bot$ is determined.)  Scaling by negative numbers is
also possible with swapping of bounds, as is multiplication.

@@

Comparing this conjecture to \conref{tangent-stable-fns}, we can see
that the difference between $\CM$, $\Lposet$ and $\Fam(\MeetSLat)$,
$\Fam(\LatGal)$ is that the latter have a separation of points from
tangents, somewhat analogous to the situation with manifolds. If this
conjecture holds, then contrary to the $\Fam(\FinVect)$ representation
of manifolds and differentiable maps, we do not throw away information
about higher derivatives. It is retained in the order structure of the
tangent fibres.
